---
title: "Lab Work Journal"
author: "BRIAN KOIMA - S84922"
date: "2024-10-16"
output:
  pdf_document: default
---

# Lab 1: Vectors, Matrices & Data Frames

## Vector Creation and Subsetting

```{r}
# Create vectors of different data types
char_list <- character(length = 0) # Empty character vector
num_list <- numeric(length = 10) # Numeric vector with default value 0
log_list <- logical(length = 3) # Logical vector with default value FALSE

# Create a logical vector with TRUE and FALSE values
log_list_2 <- c(TRUE, FALSE, FALSE, TRUE, TRUE, TRUE)

# Subsetting vectors
char_list_2 <- c("a", "b", "c", "d", "e") # Create a character vector for subsetting
char_list_2[5] # Select single element
log_list_2[2:4] # Select a range of elements
num_list[3:length(num_list)] # Select elements from a specific position to the end
```

## Vector Ordering and Concatenation

```{r}
# Vector ordering
test <- c(5, 2, 8, 1, 3) # Create a vector for sorting
sort(test, decreasing = TRUE) # Sort a vector in descending order
test[order(test, decreasing = TRUE)] # Another way to sort a vector

# Concatenate vectors
num_list_2 <- c(1, 2, 3, 4, 5)
new_num_vect <- c(num_list, num_list_2) # Combine two vectors
new_combo_vect <- c(as.numeric(num_list), log_list) # Combine vectors of different data types (convert to numeric for compatibility)
```

## Matrix Creation and Exploration

```{r}
# Create a matrix
matr <- matrix(data = c(1, 3, 5, 7, NA, 11), nrow = 2, ncol = 3) # Create a 2x3 matrix

# Explore matrix attributes
class(matr) # Check the class of the matrix
typeof(matr) # Check the data type of the matrix elements
```

## Data Frame Manipulation with `mtcars`

```{r}
# Calculate statistics from the 'mtcars' data frame
mean(mtcars$mpg) # Calculate the mean of the 'mpg' column
median(mtcars$cyl) # Calculate the median of the 'cyl' column

# Check if a row of the data frame is a list
is.list(mtcars[1, ])

# Manipulate the data frame
mtcars[c(1, 3)] # Keep the 1st and 3rd columns
mtcars[-c(1, 3)] # Remove the 1st and 3rd columns
mtcars[!duplicated(mtcars$mpg), ] # Remove duplicate rows based on the 'mpg' column
subset(mtcars, mpg < 19) # Subset rows based on a condition
mtcars[mtcars$mpg < 19, ] # Another way to subset rows
mtcars[which(mtcars$mpg < 19), ] # Another way to subset rows
mtcars[mtcars$mpg > 20 & mtcars$am == 1, ] # Subset rows based on multiple conditions
mtcars[grep("Merc", row.names(mtcars), value = TRUE), ] # Filter rows based on a pattern match
```

## Data Frame Creation and Combination

```{r}
# Create new data frames based on conditions
low_mpg <- mtcars[mtcars$mpg < 15, ]
high_mpg <- mtcars[mtcars$mpg >= 15, ]

# Combine two data frames vertically
mpg_join <- rbind(low_mpg, high_mpg)
mpg_join
# Create a data frame with random values
car_condition <- data.frame(condition = sample(c("old", "new"), replace = TRUE, size = 32))

# Combine two data frames horizontally
mpg_join <- cbind(mpg_join, car_condition)
mpg_join
```

# Vector Tasks

## Vector Arithmetic and Logical Operations

```{r}
# Create two numeric vectors
a <- c(1, 2, 3, 4, 5)
b <- c(5, 4, 3, 2, 1)

# Perform element-wise addition, subtraction, multiplication, and division
addition <- a + b
subtraction <- a - b
multiplication <- a * b
division <- a / b

# Create a logical vector indicating which elements of 'a' are greater than the corresponding elements of 'b'
greater_than <- a > b

# Print the results
print(addition)
print(subtraction)
print(multiplication)
print(division)
print(greater_than)
```

## Indexing and Subsetting

```{r}
# Create a numeric vector with 15 random integers between 1 and 100
random_vector <- sample(1:100, 15)

# Extract the elements at positions 3, 7, and 12
extracted_elements <- random_vector[c(3, 7, 12)]

# Create a new vector that contains only the even numbers from the original vector
even_numbers <- random_vector[random_vector %% 2 == 0]

# Print the results
print(random_vector)
print(extracted_elements)
print(even_numbers)
```

# Lab 2: Lists & Functions

## Creating and Manipulating Lists

```{r}
# Create a list containing different data types
my_list <- list(numbers = 1:5, fruits = c("apple", "banana", "orange"), is_sunny = c(TRUE, FALSE))

# Access elements in the list
print(my_list$numbers)  # Access using name
print(my_list[[2]])  # Access using index

# Add a new element
my_list$new_element <- c("hello", "world")

# Remove an element
my_list$is_sunny <- NULL

# Convert list to vector
new_vector <- unlist(my_list)
print(new_vector)
```

## Writing and Calling Functions

```{r}
# Define a function to add two numbers
add_numbers <- function(x, y) {
  return(x + y)
}

# Call the function
result <- add_numbers(10, 5)
print(result)
```

## Applying Functions to List Elements

```{r}
# Create a list of numeric values
number_list <- list(1, 2, 3, 4, 5)

# Define a function to double a number
double_values <- function(x) {
  return(x * 2)
}

# Apply the function to each element of the list
doubled_list <- lapply(number_list, double_values)
print(doubled_list)
```

## Applying Functions to Matrices

```{r}
# Create a matrix
my_matrix <- matrix(c(1, 2, 3, 4, 5, 6), nrow = 2, ncol = 3)

# Apply a function to rows or columns of a matrix
row_sums <- apply(my_matrix, 1, sum)
col_sums <- apply(my_matrix, 2, sum)

# Print the results
print(row_sums)
print(col_sums)
```

# Practice 1: Creating and Accessing Lists

```{r}
# Create a list containing a numeric vector, a character vector, and a logical vector
my_list <- list(numeric_vector = c(1, 2, 3),
                character_vector = c("a", "b", "c"),
                logical_vector = c(TRUE, FALSE, TRUE))

# Access and print each element in the list
print(my_list$numeric_vector)
print(my_list$character_vector)
print(my_list$logical_vector)
```

# Practice 2: Writing and Using Simple Functions

```{r}
# Write a function that adds two numbers
add_numbers <- function(x, y) {
  return(x + y)
}

# Use the function to add 5 and 3, then print the result
result <- add_numbers(5, 3)
print(result)
```

# Practice 3: Applying a Function to List Elements

```{r}
# Write a function that doubles a number
double_number <- function(x) {
  return(x * 2)
}

# Create a list of numeric values
number_list <- list(1, 2, 3, 4, 5)

# Use the function to double each value in the list and print the results
doubled_list <- lapply(number_list, double_number)
print(doubled_list)
```

# Lab 3: Lists & Functions II

## Working with Lists in R

```{r}
# Example vectors and data frame
num_vct <- c(2:5) # Numeric vector
chr_vct <- c("data", "science") # Character vector
data_frame <- data.frame(x = c("a", "b", "c", "d"), y = c(1:4)) # Simple data frame

# Creating a list
lista <- list(data_frame, num_vct, chr_vct) # List with a data frame, num_vct, and chr_vct
print(lista) # Print the list

# List structure and manipulation
str(lista) # Structure of the list
length(lista) # Length of the list
print(as.list(chr_vct)) # Convert a vector to a list

# Assigning names to the list elements
names(lista) <- c("data", "numbers", "words")
print(lista[3]) # Access 3rd element using single brackets
print(lista[[3]]) # Access 3rd element using double brackets

print(is.list(lista[3])) # Check if it is a list
print(is.list(lista[[3]])) # Check if this is also a list
print(class(lista[[3]])) # Check the class of the element

# More subsetting and list manipulation
print(lista$words) # Access an element by name
print(lista[["words"]]) # Another way to access the element
print(lista[["words"]][1]) # Access a specific value inside the list element
lista$new_elem <- c(TRUE, FALSE, FALSE, TRUE) # Add a new element to the list
print(length(lista)) # Check the length after adding the element
lista$new_elem <- NULL # Remove the new element
new_vect <- unlist(lista) # Unlist the list to create a vector
print(new_vect)
```

## Working with Functions in R

```{r}
# Simple function definition
fun <- function(x) x + 10
print(fun(5)) # Test the function

# Function with two arguments
fun2 <- function(x, y) x + y
print(fun2(3, 4)) # Test the function with two arguments

# Function with multiple expressions inside
fun <- function(x, y) {
  a <- sum(x)
  b <- sum(y)
  a - b
}
print(fun(c(5, 4, 3), c(1, 1, 1))) # Test the function
```

## Functional Programming and Operators

```{r}
print("^"(2, 2))
print(2^2)
print("+"(2, 2))
print(2 + 2)
print("-"("+"(3, 5), 2))
print("&"(">"(3, 2), TRUE))
```

## Apply Functions in R

```{r}
# Create a list and apply a function to each element
aList <- list(c(1, 2, 3), c(4, 5, 6), c(7, 8, 9), c(10, 11, 12))
res <- lapply(aList, function(x) { x[1] + x[2] - x[3] })
print(unlist(res)) # Unlist the result
```

## Working with Matrices and Apply Functions

```{r}
# Create a matrix
myMat <- matrix(c(1, 2, 3, 4, 5, 6, 7, 8, 9), nrow=3, ncol=3)

# Sum of rows
rsMyMat <- apply(myMat, 1, function(x) { sum(x) })
print(rsMyMat)

# Sum of columns
csMyMat <- apply(myMat, 2, function(x) { sum(x) })
print(csMyMat)

# Using existing functions for row and column sums
rsMyMat1 <- apply(myMat, 1, sum)
print(rsMyMat1)
csMyMat1 <- apply(myMat, 2, sum)
print(csMyMat1)

print(rowSums(myMat)) # Another way to sum rows
print(colSums(myMat)) # Another way to sum columns
```

## Task: Create a Function

```{r}
# Define a function to add two numbers
add_two_numbers <- function(a, b) {
  return(a + b)
}

# Test the function
print(add_two_numbers(4, 6)) # Should return 10
print(add_two_numbers(3, 5)) # Should return 8
print(add_two_numbers(-2, 7)) # Should return 5
```

# Lab 4: Data Visualization Using R

## Install and Load Necessary Packages

```{r}
# Install necessary packages
install.packages("ggplot2")
install.packages("gridExtra")

# Load libraries
library(ggplot2)
library(gridExtra)
```

## Scatter Plot: Horsepower vs. MPG

```{r}
p1 <- ggplot(mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  labs(title = "Scatter Plot of Horsepower vs. MPG",
       x = "Horsepower",
       y = "Miles per Gallon")
print(p1)
```

## Bar Plot: Cylinder Counts

```{r}
p2 <- ggplot(mtcars, aes(x = factor(cyl))) +
  geom_bar() +
  labs(title = "Bar Plot of Cylinder Counts",
       x = "Number of Cylinders",
       y = "Count")
print(p2)
```

## Scatter Plot with Color: Horsepower vs. MPG by Cylinder

```{r}
p3 <- ggplot(mtcars, aes(x = hp, y = mpg, color = factor(cyl))) +
  geom_point() +
  labs(title = "Horsepower vs. MPG with Cylinders Colored",
       x = "Horsepower",
       y = "Miles per Gallon",
       color = "Number of Cylinders")
print(p3)
```

## Bar Plot with Labels: Cylinder Counts

```{r}
p4 <- ggplot(mtcars, aes(x = factor(cyl), fill = factor(cyl))) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(title = "Bar Plot of Cylinder Counts with Labels",
       x = "Number of Cylinders",
       y = "Count")
print(p4)
```

## Histogram: Miles per Gallon

```{r}
p5 <- ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(title = "Histogram of Miles per Gallon",
       x = "Miles per Gallon",
       y = "Frequency")
print(p5)
```

## Box Plot: MPG by Cylinder Count

```{r}
p6 <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_boxplot() +
  labs(title = "Box Plot of MPG by Cylinder Count",
       x = "Number of Cylinders",
       y = "Miles per Gallon")
print(p6)
```

## Violin Plot: MPG by Cylinder Count

```{r}
p7 <- ggplot(mtcars, aes(x = factor(cyl), y = mpg)) +
  geom_violin() +
  labs(title = "Violin Plot of MPG by Cylinder Count",
       x = "Number of Cylinders",
       y = "Miles per Gallon")
print(p7)
```

## Line Plot: Weight vs. MPG

```{r}
p8 <- ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_line() +
  labs(title = "Line Plot of Weight vs MPG",
       x = "Weight",
       y = "Miles per Gallon")
print(p8)
```

## Density Plot: Miles per Gallon

```{r}
p9 <- ggplot(mtcars, aes(x = mpg)) +
  geom_density(fill = "lightblue") +
  labs(title = "Density Plot of Miles per Gallon",
       x = "Miles per Gallon",
       y = "Density")
print(p9)
```

## Combine All Plots into a Grid

```{r}
grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3, nrow = 3)
```

#Lab 5: Predicting Heart Disease using R

## Introduction

In this lab, we will use a dataset of heart disease cases to build predictive models. We will start by exploring the data, visualizing important relationships, and then building and evaluating several machine learning models including logistic regression, Support Vector Machine (SVM), and Random Forest.

## Part 1: Data Exploration and Visualization

```{r}
# Install required libraries
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("corrplot")) install.packages("corrplot")
library(tidyverse)
library(corrplot)
```

```{r}

# Load the heart disease dataset from URL
data_url <- 'https://raw.githubusercontent.com/koimabrian/Datasets/refs/heads/main/heart.csv'
data <- read.csv(data_url, comment = "#")

# Display the first and last few rows
head(data)
tail(data)

```

```{r}
# Get a brief overview of the data structure
glimpse(data)

```

```{r}
# Determine number of rows, columns, and column names
ncol(data)
nrow(data)
colnames(data)

# Summarize the data (descriptive statistics)
summary(data)
```

```{r}

# Create a transformed dataset with more descriptive variable labels
data2 <- data %>%
  mutate(
    sex = ifelse(sex == 1, "MALE", "FEMALE"),
    fbs = ifelse(fbs == 1, ">120", "<=120"),
    exang = ifelse(exang == 1, "YES", "NO"),
    cp = ifelse(cp == 1, "ATYPICAL ANGINA", ifelse(cp == 2, "NON-ANGINAL PAIN", "ASYMPTOMATIC")),
    restecg = ifelse(restecg == 0, "NORMAL", ifelse(restecg == 1, "ABNORMALITY", "PROBABLE OR DEFINITE")),
    slope = as.factor(slope),
    ca = as.factor(ca),
    thal = as.factor(thal),
    target = ifelse(target == 1, "YES", "NO")
  ) %>%
  mutate_if(is.character, as.factor) %>%
  dplyr::select(target, sex, fbs, exang, cp, restecg, slope, ca, thal, everything())
```

```{r}

# Bar plot to visualize the distribution of heart disease presence/absence
ggplot(data2, aes(x = target)) +
  geom_bar() +
  xlab("Heart Disease") +
  ylab("Count") +
  ggtitle("Presence & Absence of Heart Disease") +
  scale_fill_discrete(name = 'Heart Disease', labels = c("Absence", "Presence"))

```

```{r}
# Proportion of individuals with and without heart disease
prop.table(table(data2$target))

# Analyze age distribution
data2 %>%
  group_by(age) %>%
  count() %>%
  filter(n > 10) %>%
  ggplot() +
  geom_col(aes(age, n), fill = 'green') +
  ggtitle("Age Analysis") +
  xlab("Age") +
  ylab("Count")

# Compare blood pressure (trestbps) across chest pain types (cp)
data2 %>%
  ggplot(aes(x = sex, y = trestbps)) +
  geom_boxplot(fill = 'purple') +
  xlab('Sex') +
  ylab('Blood Pressure') +
  facet_grid(~ cp)


# Compare cholesterol (chol) across chest pain types (cp)
data2 %>%
  ggplot(aes(x = sex, y = chol)) +
  geom_boxplot(fill = 'orange') +
  xlab('Sex') +
  ylab('Cholesterol') +
  facet_grid(~ cp)
```

```{r}

# Create a correlation matrix
cor_heart <- cor(data2[, 10:14])

corrplot(cor_heart, method = 'square', type = 'upper')
```

## Part 2: Model Building and Evaluation

```{r, message=FALSE, warning=FALSE}
# Install required libraries
if (!require("caret")) install.packages("caret")
if (!require("pROC")) install.packages("pROC")
if (!require("e1071")) install.packages("e1071", dependencies = TRUE)
if (!require("kernlab")) install.packages("kernlab")
if (!require("randomForest")) install.packages("randomForest")
library(caret)
library(pROC)
library(e1071)
library(kernlab)
library(randomForest)
```

### Check for Missing Values and Set Random Seed

```{r check-missing-values, message=FALSE, warning=FALSE}
# Check for missing values
sum(is.na(data2))

# Set a random seed
set.seed(10)
```

### Split Data into Training and Testing Sets

```{r split-data, message=FALSE, warning=FALSE}
# Split the data into training (70%) and testing (30%) sets
inTrainRows <- createDataPartition(data2$target, p = 0.7, list = FALSE)
trainData <- data2[inTrainRows, ]
testData <- data2[-inTrainRows, ]

# Verify split ratio
cat("Ratio of training to testing data:", nrow(trainData) / (nrow(testData) + nrow(trainData)), "\n")
```

### Initialize Model Metric Storage

```{r initialize-metrics, message=FALSE, warning=FALSE}
# Initialize lists to store model metrics
AUC <- list()
Accuracy <- list()
```

### Logistic Regression Model

```{r logistic-regression, message=FALSE, warning=FALSE}
# Logistic Regression Model
logRegModel <- train(target ~ ., data = trainData, method = 'glm', family = 'binomial')
logRegPrediction <- predict(logRegModel, testData)
logRegPredictionprob <- predict(logRegModel, testData, type = 'prob')[2]

# Confusion Matrix for Logistic Regression
logRegConfMat <- confusionMatrix(logRegPrediction, testData[, "target"])

# Calculate AUC and Accuracy for Logistic Regression
AUC$logReg <- roc(as.numeric(testData$target), as.numeric(as.matrix((logRegPredictionprob))))
Accuracy$logReg <- logRegConfMat$overall['Accuracy']
```

### Support Vector Machine (SVM) Model

```{r svm-model, message=FALSE, warning=FALSE}
# SVM Model with 10-fold cross-validation
fitControl <- trainControl(method = "repeatedcv", number = 10, repeats = 10, classProbs = TRUE, summaryFunction = twoClassSummary)
svmModel <- train(target ~ ., data = trainData, method = "svmRadial", trControl = fitControl, preProcess = c("center", "scale"), tuneLength = 8, metric = "ROC")

# Predictions for SVM
svmPrediction <- predict(svmModel, testData)
svmPredictionprob <- predict(svmModel, testData, type = 'prob')[2]

# Confusion Matrix for SVM
svmConfMat <- confusionMatrix(svmPrediction, testData[, "target"])

# Calculate AUC and Accuracy for SVM
AUC$svm <- roc(as.numeric(testData$target), as.numeric(as.matrix((svmPredictionprob))))
Accuracy$svm <- svmConfMat$overall['Accuracy']
```

### Random Forest Model

```{r random-forest-model, message=FALSE, warning=FALSE}
# Random Forest Model
RFModel <- randomForest(target ~ ., data = trainData, importance = TRUE, ntree = 200)

# Predictions for Random Forest
RFPrediction <- predict(RFModel, testData)
RFPredictionprob <- predict(RFModel, testData, type = "prob")[, 2]

# Confusion Matrix for Random Forest
RFConfMat <- confusionMatrix(RFPrediction, testData[, "target"])

# Calculate AUC and Accuracy for Random Forest
AUC$RF <- roc(as.numeric(testData$target), as.numeric(as.matrix((RFPredictionprob))))
Accuracy$RF <- RFConfMat$overall['Accuracy']
```

### Compare Model Metrics

```{r compare-models, message=FALSE, warning=FALSE}
# Compare AUC and Accuracy of all models
cbind(as.data.frame(matrix(c(AUC, Accuracy), nrow = 3, ncol = 2, dimnames = list(names(AUC), c("AUC", "Accuracy")))))
```

### Display Confusion Matrices

```{r display-confusion-matrices, message=FALSE, warning=FALSE}
# Confusion matrices for all models
logRegConfMat
svmConfMat
RFConfMat
```

## Conclusion

In this lab, we explored the heart disease dataset and built three predictive models: logistic regression, Support Vector Machine (SVM), and Random Forest. The models were evaluated using AUC and accuracy metrics, with the results compared across models.

# Lab 6: Clustering using K-means (Wine Dataset Clustering Analysis)

## Load Dataset

The necessary libraries and the wine dataset are loaded.

```{r load-data}
library(tidyverse)

data_url <- 'https://raw.githubusercontent.com/koimabrian/Datasets/refs/heads/main/Wine.csv'
Wine <- read.csv(data_url, comment = "#")  # Handle potential comments in the CSV
head(Wine)
```

## Standardize Continuous Variables

The continuous variables (excluding observation number and country) are scaled.

```{r standardize-data}
data <- Wine %>% select(-Obs, -Country) %>% scale()
```

## Create Clusters with K-Means

The k-means algorithm is applied to the standardized data, setting the number of clusters to 3.

```{r kmeans-clustering}
kmeans_result <- kmeans(data, centers = 3, iter.max = 100, nstart = 100)
kmeans_result
```

## Determine the Optimal Number of Clusters

To identify the optimal number of clusters, the within-cluster sum of squares (WSS), silhouette method, and gap statistic are visualized.

```{r optimal-cluster-number, warning=FALSE, message=FALSE}
library(factoextra)
# WSS plot
fviz_nbclust(data, kmeans, method = "wss")

# Silhouette plot
fviz_nbclust(data, kmeans, method = "silhouette")

# Gap statistic
fviz_nbclust(data, kmeans, method = "gap_stat")
```

## Cluster Visualization

Clusters are visualized in a biplot, and then plotted using the original variables (Rating vs Price).

```{r cluster-biplot, warning=FALSE, message=FALSE}
# Cluster biplot
fviz_cluster(kmeans(data, centers = 3, iter.max = 100, nstart = 100), data = data)
```

### Clusters with Original Variables

Cluster labels are added to the original dataset, and clusters are visualized using the Rating and Price variables.

```{r original-variable-clusters, warning=FALSE, message=FALSE}
clusters <- kmeans(data, centers = 3, iter.max = 100, nstart = 100)
Wine <- Wine |> mutate(cluster = clusters$cluster)

# Plot clusters using Rating and Price
Wine |> ggplot(aes(x = Rating, y = Price, col = as.factor(cluster))) + 
  geom_point() + 
  labs(title = "Wine Clusters by Rating and Price", color = "Cluster")
```

## Conclusion

The k-means clustering analysis successfully grouped the wine dataset into 3 distinct clusters, which were visualized using the `Rating` and `Price` variables.
